### 2021-03-14 newvariant.tex
done for now?

Characterizing epidemic dynamics of new SARS-CoV-2 variants is critical to predicting the future course of the pandemic.
Here, we explore how assumptions about the relative strength (i.e., ratio of reproduction numbers $\RR$) and relative speed (i.e., differences in growth rates $r$) of the new variant affect conclusions about expected rates of variant takeover and viral spread under various conditions.
We show that failing to account for the possibility that the new variant has longer generation intervals can lead to considerable biases in the inferences about the relative reproductive number (strength) of the new variant.
For example, the plausible assumption that the mean generation interval of the new variant is 40\% longer than that of the wild type can lead to biases of up to 40\% in estimates of  the relative transmissibility of the new variant.
We suggest that better estimates of viral generation intervals could improve forecasts of variant takeover and viral spread.

@@

\begin{figure}[!th]
\includegraphics[width=\textwidth]{relstrength.pdf}
\caption{
\textbf{Relative strength of the new variant given observed relative speed.}
(A) Estimated relative strength of the new variant $\hat{\theta}$ given speed of the wild type
$r_w$ assuming identical generation-interval distributions.
(B) True relative strength of the new variant $\theta$ given speed of the wild type $r_w$ accounting for longer generation intervals of the new variant.
(C) Ratio between the true and the estimated relative strength of the new variant $\theta/\hat{\theta}$.
True relative strength of the new variant $\theta$ is calculated using $\delta=0.1\,\textrm{days}$, $\bar{G}_w = 5\,\textrm{days}$, and $\bar{G}_v = 7\,\textrm{days}$.
}
\label{fig:relstrength}
\end{figure}

@@

While relative speed is easier to measure, we often do not expect $\delta$ to be constant---this can depend on epidemic and demographic factors and can vary across time and space.
Instead, biological mechanisms that cause new variant to have higher strength better translate to assuming a constant value of relative strength.
For example, if the new variant has higher transmission rate,

Instead, assuming a constant relative strength $\theta$ better matches biological hypotheses about the new variant (e.g., increased transmission and immune escape).
Several studies have tried to predict the future spread of the new variant, including when it will replace the wild type (therefore, its relative speed $\delta$).
As our baseline scenario we assume $\theta = 1.8$ and ask how assumptions about the generation-interval distribution affect predictions of $\delta$---while this is considerably higher than most previous estimates of around 1.5 \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}, we chose this value as it better reflects the possibility that the new variant may have longer generation intervals (\fref{relstrength}).

\begin{figure}[!th]
\includegraphics[width=\textwidth]{relspeed.pdf}
\caption{
\textbf{Relative speed of the new variant given assumed relative strength.}
(A) Estimated relative speed of the new variant $\hat{\delta}$ given strength of the wild type $\RR_w$ assuming identical generation-interval distributions.
(B) True relative speed of the new variant $\delta$ given strength of the wild type $\RR_w$ accounting for longer generation intervals of the new variant.
(C) Difference between the true and the estimated relative speed of the new variant $\delta - \hat{\delta}$.
True relative speed of the new variant $\delta$ is calculated using $\theta=1.8$, $\bar{G}_w = 5\,\textrm{days}$, and $\bar{G}_v = 7\,\textrm{days}$.
}
\label{fig:relspeed}
\end{figure}

Like others, we first ignore possible differences between the generation-interval distributions of the wild type and the new variant and try to estimate $\delta$:
\begin{equation}
\hat{\delta} = \frac{(\theta \RR_w)^{\kappa_w} - \RR_w^{\kappa_w}}{\kappa_w \bar{G}_w}.
\end{equation}
In this case, the estimated $\hat{\delta}$ increases as the underlying strength of the wild type $\RR_w$ increases (\fref{relspeed}A).
Assuming a narrower distribution causes $\hat{\delta}$ to be less sensitive to $\RR_w$;
when $\kappa = 0$ (delta-distributed generation intervals), $\hat{\delta}$ remains constant.

When we allow the mean generation interval of the new variant to be long ($\bar{G}_v > \bar{G}_w$), the true $\theta$--$\delta$ relationship is given by:
\begin{equation}
\delta = \frac{(\theta \RR_w)^{\kappa} - 1}{\kappa \bar{G}_v} - \frac{\RR_w^{\kappa} - 1}{\kappa \bar{G}_w}.
\end{equation}
In this case, the direction of the effect of $\RR_w$ on $\delta$ can change depending on the value of $\kappa$.
As $\kappa \to 1$, we have
\begin{equation}
\delta \to \frac{(\bar{G}_w \theta - \bar{G}_v) \RR_w + \bar{G}_v - \bar{G}_w}{\bar{G}_v \bar{G}_w},
\end{equation}
which changes linearly with $\RR_w$ with a slope of $(\bar{G}_w \theta - \bar{G}_v)/\bar{G}_v \bar{G}_w$ (which can be either negative or positive).
For the values we consider, this increases with $\RR_w$ (\fref{relspeed}B).
On the other hand, as $\kappa \to 0$, we have
\begin{equation}
\delta \to \frac{\log(\theta \RR_w)}{\bar{G}_v} - \frac{\log(\RR_w)}{\bar{G}_w},
\end{equation}
which always decreases with $\RR_w$ provided that $\bar{G}_v > \bar{G}_w$ (\fref{relspeed}B).
Overall, estimates of $\delta$ become less sensitive to values of $\RR_w$ and range between 0.05/day and 0.15/day, which is consistent with the observed values of $\delta$ from data.
Therefore, assuming equal generation intervals generally leads to over-estimation of $\delta$ (\fref{relspeed}C).

>>>>

@@

\begin{figure}[!th]
\includegraphics[width=\textwidth]{relstrength.pdf}
\caption{
\textbf{Relative strength of the new variant given observed relative speed.}
(A) Estimated relative strength of the new variant $\hat{\theta}$ given speed of the wild type
$r_w$ assuming identical generation-interval distributions.
(B) True relative strength of the new variant $\theta$ given speed of the wild type $r_w$ and ratio between the mean generation interval of the new variant $\bar{G}_v$ and that of the wild type $\bar{G}_w$.
(C) Ratio between the true and the estimated relative strength of the new variant $\theta/\hat{\theta}$.
True relative strength of the new variant $\theta$ is calculated using $\delta=0.1\,\textrm{days}$, $\bar{G}_w = 5\,\textrm{days}$, and $\kappa = 1/5$.
}
\label{fig:relstrength}
\end{figure}

@@

While relative speed is easier to measure, we often do not expect $\delta$ to be constant---this can depend on epidemic and demographic factors and can vary across time and space.
Instead, biological mechanisms that cause new variant to have higher strength better translate to assuming a constant value of relative strength $\theta$.
For example, if the new variant has a higher transmission rate, the ratio between the strength of the new variant and that of the wild type will remain constant ($=\theta$) under strength-like interventions that reduces the strength of both strains by a constant amount.
However, this is not necessarily true under speed-like interventions (e.g., contact tracing and self-isolation), which depends on time since infection: if the generation-interval distributions are different, the effectiveness of interventions will also differ.
For now, we consider consequences of assuming a constant relative strength $\theta$ and predicting the relative speed $\delta$---an assumption often made (either implicitly or explicitly) in studies that have tried to predict the future spread of the new variant, including when it will replace the wild type.
As our baseline scenario we assume $\theta = 1.5$.

\begin{figure}[!th]
\includegraphics[width=\textwidth]{relspeed.pdf}
\caption{
\textbf{Relative speed of the new variant given assumed relative strength.}
(A) Estimated relative speed of the new variant $\hat{\delta}$ given strength of the wild type $\RR_w$ assuming identical generation-interval distributions.
(B) True relative strength of the new variant $\delta$ given strength of the wild type $\RR_w$ and ratio between the mean generation interval of the new variant $\bar{G}_v$ and that of the wild type $\bar{G}_w$..
(C) Difference between the true and the estimated relative speed of the new variant $\delta - \hat{\delta}$.
True relative speed of the new variant $\delta$ is calculated using $\theta=1.5$, $\bar{G}_w = 5\,\textrm{days}$, and $\kappa=1/5$.
}
\label{fig:relspeed}
\end{figure}

Like others, we first ignore possible differences between the generation-interval distributions of the wild type and the new variant and try to estimate $\delta$:
\begin{equation}
\hat{\delta} = \frac{(\theta \RR_w)^{\kappa} - \RR_w^{\kappa}}{\kappa \bar{G}_w}.
\end{equation}
Here, we see that the estimated $\hat{\delta}$ increases as the underlying strength of the wild type $\RR_w$ increases (\fref{relspeed}A).
In other words, we expect the relative speed of the new variant to change as epidemiological conditions change, suggesting that a logistic model that assumes constant growth rate may not be appropriate for modeling the proportion of the new variant.
A narrower distribution causes $\hat{\delta}$ to be less sensitive to $\RR_w$;
when $\kappa = 0$ (delta-distributed generation intervals), $\hat{\delta}$ remains constant.

When we allow the mean generation interval of the new variant to be different ($\bar{G}_v \neq \bar{G}_w$), the true $\theta$--$\delta$ relationship under gamma approximation is given by \citep{park2019practical}:
\begin{equation}
\delta = \frac{(\theta \RR_w)^{\kappa} - 1}{\kappa \bar{G}_v} - \frac{\RR_w^{\kappa} - 1}{\kappa \bar{G}_w}.
\end{equation}
In this case, $\delta$ generally decreases as the mean generation interval of the new variant increases, except when both $\RR_v < 1$ and $\RR_w < 1$ (\fref{relspeed}B).
In some extreme cases (e.g., when $\bar{G}_v$ and $\RR_v$ are very low or when $\bar{G}_v$ and $\RR_v$ are very large), we can obtain $\delta < 0$---an unlikely scenario given that fast replacement of the new variant have been observed under mixed conditions.
This suggests that differences in generation intervals between the new variant and the wild type are likely to be not too big.
Finally, assuming equal generation intervals generally leads to both under- and over-estimation of $\delta$ (\fref{relspeed}C).

\section{Inferring relative strength from incidence data, in theory}

In practice, epidemiological conditions change over time due to intervention measures and susceptible depletion effects.
Such changes can be characterized by ``time-varying'' or ``instantaneous'' reproduction numbers, $\RR(t)$ \citep{fraser2007estimating}.
Assuming that the generation-interval distribution does not change over time, the instantaneous reproduction numbers of the new variant and of the wild type can be estimated from their corresponding incidence curves---an approach popularized by \cite{cori2013new}:
\begin{equation}
\RR_x(t) = \frac{i_x(t)}{\int_0^\infty i_x(t-\tau) g_x(\tau) \dtau}.
\end{equation}
Under strength-like intervention measures that reduce transmission rates by a constant amount, we expect ratios between true reproduction numbers to remain constant and correspond to the true relative strength: $\RR_v(t)/\RR_w(t) = \theta$.
However, if the assumed generation-interval distribution $\hat{g}(\tau)$ is different from the true distribution, then the ratio between the estimated reproduction numbers $\hat{\theta}(t) = \hat{\RR}_v(t)/\hat{\RR}_w(t)$ need not be constant, where
\begin{equation}
\hat{\RR}_x(t) = \frac{i_x(t)}{\int_0^\infty i_x(t-\tau) \hat{g}(\tau) \dtau}.
\end{equation}

Here, we simulate three different scenarios using a two strain renewal equation that assumes perfect cross-immunity (\fref{Rtbias}): 
(1) the wild type and the variant have the same generation-interval distributions, which match the assumed distribution (\fref{Rtbias}A--D);
(2) the assumed distribution matches the generation-interval distribution of the wild type but has a longer mean than the mean generation interval of the variant (\fref{Rtbias}E--H); and
(3) the assumed distribution matches the generation-interval distribution of the wild type but has a shorter mean than the mean generation interval of the variant (\fref{Rtbias}I--L).
Then, we compare the estimated ratio $\hat{\theta}(t) = \hat{\RR}_v(t)/\hat{\RR}_w(t)$ with the true ratio $\theta = \RR_v(t)/\RR_w(t)$.
For simplicity, we did not model non-pharmaceutical interventions;
instead, we let epidemic dynamics to be determined by susceptible depletion alone.

\begin{figure}[!th]
\includegraphics[width=\textwidth]{Rtbias.pdf}
\caption{
\textbf{Estimates of relative strength over time under different scenarios.}
(A, E, I) Daily incidence of infection caused by the new variant and the wild type.
(B, F, J) True (solid lines) and estimated (dashed lines) reproduction numbers of the new variant and the wild type over time.
(C, G, K) True (purple) and estimated (orange) ratios between reproduction numbers of the new variant and the wild type over time.
(D, H, L) Relationship between estimated reproduction numbers on a log-log sacle.
The dashed lines represent the one-to-one line. 
The dotted lines represent the $\RR_v=\theta \RR_w$ line. 
Top row: the assumed mean generation-interval (5 days) is equal to the mean generation-interval of the wild type and the variant.
Middle row: the assumed mean generation-interval (5 days) is equal to the mean generation-interval of the wild type but is longer than that of the varant (3.5 days).
Bottom row: the assumed mean generation-interval (5 days) is equal to the mean generation-interval of the wild type but is longer than that of the varant (6.5 days).
For all simulations, squared coefficient of variation in generation intervals is assumed to equal $\kappa = 1/5$.
}
\label{fig:Rtbias}
\end{figure}

When the assumed distribution matches the true distributions, the estimated ratio remains constant (\fref{Rtbias}C).
However, if the generation-interval distribution of the variant is different from the assumed distribution, the ratio changes over time (\fref{Rtbias}G,K).
In particular, if the true generation intervals have a shorter (or longer) mean than the assumed distribution, we over-estimate (or under-estimate) $\RR_v(t)$ during the growth phase and under-estimate (or over-estimate) $\RR_v(t)$ during the decay phase (\fref{Rtbias}F,J), which further translates to biases in the estimated relative strength (\fref{Rtbias}G,K).

We can also assess these biases by plotting the estimated strength of the variant $\hat{\RR}_v(t)$ against the estimated strength of the wild type $\hat{\RR}_w(t)$ on a log-log scale (\fref{Rtbias}D,H,L).
When generation-interval distributions are correctly specified, we obtain a straight line parallel to the one-to-one line. 
In this case, the slope corresponds to the logarithm of the relative strength $\log(\theta)$ (\fref{Rtbias}D).
When the assumed mean generation interval is longer (or shorter) than the that of the variant, we over-estimate (or under-estimate) the slope (\fref{Rtbias}H,L) on a log-log plot.

\section{Inferring relative strength from case data, in practice}

Seb Funk's analysis here.
